Penalized regression for left-truncated and right-censored survival data

### 2021-01-08 manuscript.tex
add CGDB citation

\begin{abstract}
    High-dimensional data are becoming increasingly common in the medical field as large volumes of patient information are collected and processed by high-thoroughput screening, electronic health records (EHRs), and comprehensive genomic testing. Statistical models that attempt to study the effects of many predictors on survival typically implement feature selection or penalized methods to mitigate the undesirable consequences of overfitting. In some cases survival data is also left-truncated which can create an immortal time bias, but penalized survival methods that adjust for left truncation are not commonly implemented. To address these challenges, we implement a penalized Cox proportional hazards model for left-truncated and right-censored survival data and assess implications of left truncation adjustment on bias and inference. We use simulation studies and examples from high-dimensional real-world EHR and genomics data to highlight the pitfalls of failing to account for left truncation in survival modeling.
\end{abstract}

@@

\subsection{Real-world use case}
We consider the Clinico-Genomic Database (CGDB) offered jointly by Flatiron Health and Foundation Medicine. The CGDB is an oncology research platform that combines real-world, patient-level clinical data and outcomes with patient-level genomic data to provide a comprehensive patient and tumor profile. Patients in the CGDB have received at least one Foundation Medicine tumor genomic test as well as have had their electronic health records captured by Flatiron Health. Together, these data sources represent thousands of potential predictors per patient, ranging from demographic, laboratory, and treatment information to the specific mutations detected on each biomarker included in a Foundation Medicine baitset.

>>>>

\begin{abstract}
    High-dimensional data are becoming increasingly common in the medical field as large volumes of patient information are collected and processed by high-thoroughput screening, electronic health records (EHRs), and comprehensive genomic testing. Statistical models that attempt to study the effects of many predictors on survival typically implement feature selection or penalized methods to mitigate the undesirable consequences of overfitting. In some cases survival data is also left-truncated which can create an immortal time bias, but penalized survival methods that adjust for left truncation are not commonly implemented. To address these challenges, we implement a penalized Cox proportional hazards model for left-truncated and right-censored survival data and assess implications of left truncation adjustment on bias and inference. We use simulation studies and a high-dimensional, real-world clinico-genomic database (CGDB) to highlight the pitfalls of failing to account for left truncation in survival modeling.
\end{abstract}

@@

\subsection{Real-world use case}
We consider the Clinico-Genomic Database (CGDB) offered jointly by Flatiron Health and Foundation Medicine \cite{singal2017cgdb}. The CGDB is a US-based oncology research platform that combines real-world, patient-level clinical data and outcomes with patient-level genomic data to provide a comprehensive patient and tumor profile. The de-identified data originated from approximately 280 US cancer clinics (~800 sites of care). Patients in the CGDB have received at least one Foundation Medicine comprehensive genomic profiling test as well as have had their electronic health records captured by Flatiron Health. The database comprises patient-level structured and unstructured data, curated via technology-enabled abstraction, linked to Foundation Medicine genomic data by de-identified, deterministic matching. The data are de-identified and subject to obligations to prevent re-identification and protect patient confidentiality. Altogether, the CGDB represents thousands of potential predictors per patient, ranging from demographic, laboratory, and treatment information to the specific mutations detected on each biomarker included in a Foundation Medicine baitset.
